---
title: "CityFibre - 4a. Data Mapping - google maps"
output:
  html_notebook:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    df_print: paged
---

<div class="alert alert-block alert-success">

**Part 4a - CityFibre google mapping**

This notebook steps through the sixth step of the CityFibre data analysis problem: using google maps to plot the location of the priority postcodes

The steps involved cover:

* Export a file with the required pin information
* Import this into google maps
* Select the marker fields

</div>


# Plot postcode pins on google maps

Write out a file with pin information

First need to set up a link to where the output files are stored
```{r}
output_file_path <- "/Users/jowatts/github/cityfibre-schools/outputs/"
```

```{r}
gmaps <- cf_pc_focus %>%
  select(city, postcode, latitude, longitude, total_score, scheme_references, nobroadband_rate, 
         simd10, pc_type, ur_class)

write_csv(gmaps, file.path(output_file_path, "cf_prioritised_pc.csv"))
```

Go to [https://www.google.com/mymaps](https://www.google.com/mymaps)

Select "Create a new map"

Give the map a suitable title

Import the exported gmaps .csv file

Select latitude and longitude as the position fields

Select Postcode as the marker title

All other data will be imported and will be visible on the marker.





