---
title: "CityFibre - 4b. Data Mapping - postcode boundaries"
output:
  html_notebook:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    df_print: paged
---

<div class="alert alert-block alert-success">

**Part 4b - CityFibre priority postcode mapping**

This notebook steps through the seventh step of the CityFibre data analysis problem: mapping the priority postcodes

The purpose of this exercise is to be able to easily share the location and information about the priority postcodes and also to visually see where they are located.

Are they all in one city?

Are in the centres, or on the edges of the cities?

The mapping activity will involve use of additional packages and an open source mapping tool called [Leaflet](https://leafletjs.com/)

The steps involved with mapping the postcodes onto a map include:

* Read in the postcode boundaries
* Convert coordinates to longitude and latitude
* Merge in any additional data
* Create pop-up labels for each postcode
* Convert to a spatial file type
* Import this information to the map
* Save the map as an html file for sharing

</div>

# Set up the mapping environment

<div class="alert alert-block alert-info">

Additional packages are required to support the mapping activities.

Ensure they are installed and loaded.

</div>

```{r}
if (!require(leaflet)){
  install.packages("leaflet")
  library(leaflet)
}

if (!require(htmlwidgets)){
  install.packages("htmlwidgets")
  library(htmlwidgets)
}
```

# Import postcode boundary files

<div class="alert alert-block alert-info">

The geometry information for the postcodes is available in the postcode shape files.

</div>

Using `st_read()`, read in the postcode boundary shape file - "PC_CUT_20_2.shp" tidying up the column names on import
```{r}
pc_geom_input <- st_read(file.path(input_file_path, "PC_CUT_20_2/PC_CUT_20_2.shp")) %>%
  clean_names()
```

Using `head()` have quick look at this data
```{r}
head(pc_geom_input)
```
It can be seen that each row is a postcode and the boundaries are contained within the geometry information. The geometry information is a set of polygons defined by points.

Subset the file to just retain only the postcode and geometry fields.
```{r}
pc_boundaries <- select(pc_geom_input, postcode, geometry)
```

# Convert coordinate system

<div class="alert alert-block alert-info">

This is similar to a dataframe, but contains geometric information in a specific coordinate system.
This file contains polygons defining the boundaries of each postcode.

The coordinate system currently used is the OSGB 1936 /British National Grid, which maps the boundaries to Eastings and Northings.  
[Ordinance Survey National Grid](https://en.wikipedia.org/wiki/Ordnance_Survey_National_Grid)

To use a mapping tool, these need to be converted to values of longitude and latitude.
This can be done using a coordinate system transformation.
The value required is 4326, which links to the [World Geodetic System](https://en.wikipedia.org/wiki/World_Geodetic_System)

</div>

Transform the geometry to longitude and latitude using `st_transform()` and setting the coordinate reference system `crs` to 4326
```{r}
pc_latlong <- st_transform(pc_boundaries, crs = 4326)
```

# Merge in additional data

Finally recreate the `postcode2` field to enable merging in of additional data items.
Use `str_to_upper()` for uppercase and `str_remove_all(varname, " ")` to remove spaces
```{r}
pc_latlong <- mutate(pc_latlong, postcode2 = str_remove_all(str_to_upper(postcode), " "))
```

Keep only the priority postcode geometry by doing a `left_join()` with the cf_pc_focus file

Geometry is only available for "small" postcodes, so also filter out those that do not have any geometry
```{r}
cf_focus_latlong <- cf_pc_focus %>%
  select(postcode2, scheme_references, ur_class, simd10, simd_rank, nobroadband_rate) %>%
  left_join(pc_latlong, by = "postcode2") %>%
  filter(!is.na(geometry))
```

# Create pop-up labels

Create a popup labels for each postcode on the map for both scoring approaches.
This is done by creating a formatted text string for each label.
```{r}
cf_focus_latlong <- cf_focus_latlong %>%
  mutate(pop_label = paste(sep = "<br/>",
                           paste0("<b>",postcode,"</b>"),
                           paste0("Scheme: ", scheme_references), 
                           paste0("Urban Rural Classification: ", ur_class), 
                           paste0("SIMD Decile: ", simd10),
                           paste0("SIMD Rank: ", simd_rank),
                           paste0("No broadband access rate: ", nobroadband_rate, "%")))
```

# Convert to a spatial file type

The file will need to be converted to a spatial file type for mapping
```{r}
cf_focus_latlong_sf <- st_as_sf(cf_focus_latlong)
```

Create coordinates for the centre of the map to be around the centre of Scotland
```{r}
centre_long <- -4.2
centre_lat <- 56.5
```

# Import into a leaflet map

Draw the boundaries using the leaflet package and the spatial file type
```{r}

map <- leaflet() %>%
  setView(lng = centre_long, lat = centre_lat, zoom = 7) %>%
  addTiles() %>%
  addPolygons(data = cf_focus_latlong_sf,
              weight = 2, 
             color = "#ff8c00", 
             stroke = TRUE, 
             fillOpacity = 0.5,
             highlight = highlightOptions(color = "#000000",
                                          weight = 2,
                                          stroke = TRUE,
                                          fillOpacity = 0.8,
                                          bringToFront = TRUE),
             popup = ~pop_label
             )

map  
```

Define the directory where the output file should be saved by defining the `output_file_path`
```{r}
output_file_path <- "/mypathtodata/outputs/"
```

Save this map to the output directory so it can be shared with others
```{r}
saveWidget(map, file = file.path(output_file_path, "cf_prioritypc.html"))
```

This `.html` file can now be viewed in a web browser and the locations of the prioritised postcodes and their information viewed on a map.

# Extension - Map experimentation

Experiment with changing the information that is displayed in the tool tips.

Reformat the popup labels.

Change the colours that are used for background and highlight on the leaflet map.

Change the dataset for the differently prioritised scoring dataset.