---
title: "CityFibre - 2c. Data Cleaning - SIMD dataset"
output:
  html_notebook:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    df_print: paged
---

<div class="alert alert-block alert-success">

**Part 2b - SIMD dataset cleaning**

This notebook steps through the third step of the CityFibre data analysis problem: cleaning the SIMD dataset.

The SIMD dataset currently contains numeric values stored as percentages that need to have the "%" removed and converted to proper numeric values.

The column names are also not intuitive, so may lead to confusion in future analysis. It is therefore necessary to rename them.

The steps involved include: 

* Selecting the required columns
* Convert to numeric those columns that are in string format
* Rename all columns to a descriptive variable name

</div>

# SIMD data cleaning

## SIMD retain columns of interest

Use the `names()` function to print out all the fields in the `simd_input` dataset:
```{r}
names(simd_input)
```

<div class="alert alter-block alert-info">

The problem statement requires the analyst to focus on SIMD, income, employment and broadband access.
The names of these fields in the dataset are:

* rankv2 - SIMD rank
* inc_rate
* emp_rate
* g_acc_brdbnd

Other fields to retain for reference or further analysis are:

* data_zone
* dz_name
* sape2017 - population
* wape2017 - working age population

Other fields may also be retained. The detail of each field can be found in the SIMD technical notes:

[SIMD technical notes](https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/09/simd-2020-technical-notes/documents/simd-2020-technical-notes/simd-2020-technical-notes/govscot%3Adocument/SIMD%2B2020%2Btechnical%2Bnotes.pdf)

</div>

Use `select()` to select the required fields. 
Use `st_drop_geometry() to convert the `sf` object into a data frame.
Name the output dataset `simd_retain`.
```{r}
simd_retain <- simd_input %>%
  select(data_zone, 
         dz_name, 
         sape2017, 
         wape2017, 
         rankv2, 
         quintilev2, 
         decilev2, 
         inc_rate, 
         emp_rate,
         g_acc_brdbnd) %>%
  st_drop_geometry()
```

## Convert string columns to numerics

Use `head()` to review the first few rows of the retained dataset
```{r}
head(simd_retain)
```
Use `summary()` to review the format of the remaining dataset
```{r}
summary(simd_retain)
```
From this it can be seen that `inc_rate`, `emp_rate` and `g_acc_brdbnd` are all character columns which contain numeric values.

<div class="alert alter-block alert-info">

To fix these columns is a two-step process:

* remove the %'s 
* convert the columns to numeric

The `str_remove(varname, "%")` function can be used to remove the %s
The `as.numeric(varname)` can then be used to change the variable type.

</div>

Use `str_remove()` and `as.numeric()` within a `mutate()` statement for each of the columns that require fixing. Write this out to a new cleaned dataset `simd_clean`
```{r}
simd_clean <- mutate(simd_retain, 
                     inc_rate = as.numeric(str_remove(inc_rate, "%")),
                     emp_rate = as.numeric(str_remove(emp_rate, "%")),
                     g_acc_brdbnd = as.numeric(str_remove(g_acc_brdbnd, "%")))
```

## Rename columns to be more descriptive

Use `names()` to write out the column names on the cleaned dataset
```{r}
names(simd_clean)
```

<div class="alert alter-block alert-info">

The column names are not very intuitive.

By referring to the Technical Notes it be seen that:

* sape2017 is total population within the datazone
* wape2017 is the working age population within the datazone
* rankv2 is the rank across the whole of Scotland for that datazone. Low ranks are the most deprived, high ranks are the least deprived.
* quintilev2 splits the ranks up into 5 groups containing 20% each. Quintile 1 is the most deprived. Quintile 5 is the least deprived.
* decilev2 splits the ranks up into 5 groups containing 10% each. Similar to quintiles, 1 is the most deprived and 10 is the least deprived.
* inc_rate is the percentage of adults in the datazone receiving income support.
* emp_rate is the percentage of adults in the datazone receiving employment support
* g_acc_brdbnd is the percentage of premises that **do not** have access to superfast broadband of 30Mb/s

</div>

For all the columns, use `rename()` to give them a variables a name that is more descriptive.
Remember to continue to use lowercase letters and put '_' between words.
```{r}
simd_clean <- simd_clean %>%
  rename(total_population = sape2017,
         working_age_population = wape2017,
         simd_rank = rankv2,
         simd5 = quintilev2,
         simd10 = decilev2,
         income_support_rate = inc_rate,
         employment_support_rate = emp_rate,
         nobroadband_rate = g_acc_brdbnd)
```

Use `head()` to review the first few rows of the output file
```{r}
head(simd_clean)
```

